Сплинты корневых систем для специальных подалгебр Ли

### 2015-03-21 AffineLieAlgebras/Splints/special-paper-ru.tex
Russian translation started

\usepackage{amsmath,amssymb,amsthm}
\usepackage{pb-diagram}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\usepackage{multicol}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[english]{babel}

@@

\title{Splints of root systems for special Lie subalgebras}

\author{V.D.~Lyakhovsky$^1$, A.A.~Nazarov$^{1,2}$, P.I.~Kakin$^{1}$\\
  {\small $^1$ Department of High-energy and elementary particle physics,}\\ {\small St Petersburg State University}\\
  {\small 198904, Saint-Petersburg, Russia,}\\
  {\small $^{2}$ e-mail: antonnaz@gmail.com}}
%\date{}
%\address{ $^1$ Department of High-energy and elementary particle physics,  St Petersburg State University, 198904, Saint-Petersburg, Russia}
%\address{ $^{2}$ e-mail: anton.nazarov@hep.phys.spbu.ru}

@@

\begin{abstract}
  Splint is a decomposition of root system into a union of root systems. Splint of root system for a
  simple Lie algebra appears naturally in the studies of (regular) embeddings of reductive
  subalgebras. Splint can be used to construct branching rules. We consider a special embedding of a
  Lie subalgebra into a simple Lie algebra. We classify the projections of algebra root systems and
  single out the conditions of splint appearance and coincidence of branching coefficients with
  weight multiplicities. While such a coincidence is not very common it is connected with
  Gelfand-Tsetlin basis.

\noindent{\it Keywords\/}: Lie algebra, special subalgebra, branching, weight multiplicity, splint
\end{abstract}

\section{Introduction}
\label{sec:introduction}

The notion of splint was introduced by David Richter in the paper \cite{richter2008splints}. Splint
is a decomposition of a root system into a disjoint union of the images of two or more embeddings of some
other root systems. An embedding $\phi$ of a root system $\Delta_1$ into a root system $\Delta$ is a
bijective map of roots of $\Delta_{1}$ to a (proper) subset of $\Delta$ that commutes with the vector
composition law in $\Delta_{1}$ and $\Delta$.
\begin{equation*}
\phi:\Delta_1 \longrightarrow \Delta
\end{equation*}
\begin{equation*}
\phi \circ (\alpha + \beta) =\phi \circ \alpha + \phi \circ \beta,
\,\,\, \alpha,\beta \in \Delta_1
\end{equation*}

Note that the image $Im(\phi)$ is not required to inherit the root system properties except the
addition rules equivalent to the addition rules in $\Delta_{1}$ (for pre-images). If an embedding
$\phi$ preserves the angles between the roots it is called ``metric''. Two embeddings $\phi_1$ and
$\phi_2$ can splinter $\Delta$ when the latter can be presented as a disjoint union of images
$Im(\phi_1)$ and $Im(\phi_2)$.

Why one would consider non angle-preserving maps of root systems? Additive properties of the root system
determine the structure of the Verma module:
    \begin{equation*}
      M^{\mu}=U(\gf)\underset{U(\bff_{+})}{\otimes} D^{\mu}(\bff_{+})=\{(E^{-\alpha_{1}})^{n_{1}}\dots (E^{-\alpha_{s}})^{n_{s}} \left|v_{\mu}\right>\}_{\alpha_{i}\in\Delta^{+}}^{n_{i}=0,1,\dots}
    \end{equation*}
Here $\bff_{+}$ is the Borel subalgebra of $\gf$, $D^{\mu}$ is its one-dimensional representation,
$\left|v_{\mu}\right>$ is the highest weight vector and $E^{-\alpha_{j}}$ are lowering operators
that are in correspondence with positive roots $\alpha_{j}\in \Delta^{+}$. 
The Weyl character formula expresses a character of an irreducible representation as a combination of
the characters of the Verma modules: 
\begin{equation*}
  \mathrm{ch} L^{\mu}=\frac{\sum_{w\in W} \varepsilon(w) e^{w(\mu+\rho)-\rho}}{\sum_{w\in W}\varepsilon(w) e^{w\rho-\rho}}=\sum_{w\in W} \varepsilon(w)\; \mathrm{ch} M^{w(\mu+\rho)-\rho}
\end{equation*}
Here the sum is over elements of the Weyl group and their actions $w\triangleright \mu
=w(\mu+\rho)-\rho$ depend on angles between roots. But we can rewrite the Weyl character formula
representing the Weyl group elements as the products of the reflections $s_{\alpha}, \alpha \in S$ in
the hyperplanes orthogonal to the simple roots: $w=s_{\alpha_{1}}\cdot s_{\alpha_{2}}\dots$. The reflections act
on the root system by permutations, so the composition of such an action with the embedding $\phi$ is easily
obtained. The orbit of the Weyl group action $w\triangleright \mu, w\in W$, can be constructed by
subtractions of roots from the highest weight $\mu$. Let's denote the Dynkin labels of $\mu$ by
$(\mu_{1},\dots \mu_{r})$. Then $\mu-\mu_{1}\alpha_{1}, \mu-\mu_{2}\alpha_{2},\dots,
\mu-\mu_{r}\alpha_{r}, \alpha_{i}\in S$, are the points of the orbit that can be obtained by
the elementary reflections $s_{\alpha_{i}}, \alpha_{i}\in S$. The next set of points that are obtained by
two consecutive reflections $s_{\alpha_{i}}\cdot s_{\alpha_{j}}\triangleright\mu$ are obtained as
the subtraction $\mu-\mu_{i}\alpha_{i}-\mu_{j} (s_{\alpha_{i}}\alpha_{j})$. Continuing this way we
construct the image of the singular element $\Psi=\sum_{w\in W} \varepsilon(w) e^{w(\mu+\rho)-\rho}$
after the embedding $\phi$ \cite{2011arXiv1111.6787L}. The multiplicities in the character of the Verma
module are unchanged by the embedding since they are determined by the additive properties of the roots.
So we see that the weight multiplicities of irreducible modules are also preserved by the embedding.

The root system of the regular subalgebra is contained in the root system of the algebra so the splint is useful in the
computation of branching coefficients.

In the paper \cite{2011arXiv1111.6787L} it was proven that the existence of splint leads to the
coincidence of branching coefficients with weight multiplicities under certain conditions. We denote a
Lie algebra by $\gf$ and consider it's subalgebra $\af$. If $\af$ is a regular subalgebra, its root
system $\Delta_{\af}$ is contained in $\Delta_{\gf}$. Branching coefficients $b^{(\mu)}_{\nu}$
appear in the decomposition of an irreducible representation $L^{\mu}_{\gf}$ of $\gf$ to the sum of
irreducible representations of $\af$:
\begin{equation}
  \label{eq:1}
  L^{\mu}_{\gf}=\bigoplus_{\nu} b^{(\mu)}_{\nu} L^{\nu}_{\af}
\end{equation}

Assume that the root system $\Delta_{\gf}$ splinters as $\Delta_{\gf}=\Delta_{\af} \cup
\phi(\Delta_{\sfr})$, where $\phi$ is an embedding of a root system $\Delta_{\sfr}$ of some semisimple
Lie algebra $\sfr$. Then branching coefficients $b^{(\mu)}_{\nu}$ for the reduction
$L^{(\mu)}_{\gf\downarrow \af}$ coincide with the weight multiplicities $m^{(\tilde \mu)}_{\nu}$ in the $\sfr$-representations provided certain technical condition holds \cite{2011arXiv1111.6787L}.
The highest weight $\tilde\mu$ of the $\sfr$-representation is calculated from the Dynkin labels of the highest
weight $\mu$ of the $\gf$-module.
The proof of the coincidence of the branching coefficients with the weight multiplicities of
$\sfr$-module is based upon the decomposition of singular element of the algebra $\gf$ into singular
elements for $\sfr$:
\begin{equation}
 \Phi ^{\mu }_{\gf} = \sum_{w\in W_{\frak{a}}}\epsilon\left( w\right)w\left( e^{\mu +\rho _{\gf}}\phi\left( e^{-\widetilde{\mu }}\Psi^{\widetilde{\mu }}_{\sfr}\right)\right),
\end{equation}
where $\epsilon\left( w\right)$ is a determinant of the element $w$ of the Weyl group $W_{\af}$ of
the subalgebra $\af$. The action of the Weyl group on the weights is extended to the algebra of formal
exponents by the rule $w\left(e^{\nu}\right)=e^{w\nu}, w\in W$. Hence, the multiplicity $M_{\left(
      \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}$ of the weight $\widetilde{\nu }$ from the
  weight diagram of the algebra $\sfr$ with the highest weight $\tilde\mu$ defines the branching
  coefficient $b_{\nu }^{(\mu )}$ for the highest weight $\nu =\left( \mu -\phi \left(
      \widetilde{\mu }-\widetilde{\nu }\right) \right) $:
\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{(\mu )}=M_{\left( \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}. 
\label{bran1}
\end{equation}

We use similar approach to study special embeddings. We consider special embeddings of Lie
subalgebras into a Lie algebra. In this case the root system of the subalgebra is not contained in
the root system of the algebra. So the original motivation for splints is not applicable. But we can
consider the projection of the root system of the algebra on the root space of the subalgebra. Such
a projection is not a root system anymore, but it satisfies milder conditions (Section
\ref{sec:spec-embedd-proj}). It is possible to classify the splints of the projection using the fact
that the adjoint representation of the algebra is decomposed into the direct sum of representations
of the subalgebra. We show that if there is a splint then branching coefficients coincide with
weight multiplicities of representations of another Lie algebra (Section
\ref{sec:splints-spec-embedd}). Thus in this case branching coefficients are very easy to compute.

The use of representation theory of the subalgebra allows us to classify all the splints for the projections
of the algebra root system. We also apply this method to metric splints and regular subalgebras and get
a unified treatment. We obtain the Gelfand-Tsetlin rules for regular and special embeddings this way.

In conclusion \ref{sec:conclusion} we discuss the cases when the projection of the root system does
not fall into the classification mentioned above.

\section{Special embeddings and projections of root system}
\label{sec:spec-embedd-proj}

>>>>

\usepackage{amsmath,amssymb,amsthm}
\usepackage{pb-diagram}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\usepackage{multicol}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}

@@

\title{Сплинты корневых систем для специальных подалгебр Ли}

\author{В.Д.~Ляховский$^1$, А.А.~Назаров$^{1,2}$, П.И.~Какинь$^{1}$\\
  {\small $^1$ Кафедра физики высоких энергий и элементарных частиц,}\\ {\small Санкт-Петербургский государственный университет}\\
  {\small 198904, Санкт-Петербург, Россия}\\
  {\small $^{2}$ e-mail: antonnaz@gmail.com}}
%\date{}
%\address{ $^1$ Department of High-energy and elementary particle physics,  St Petersburg State University, 198904, Saint-Petersburg, Russia}
%\address{ $^{2}$ e-mail: anton.nazarov@hep.phys.spbu.ru}

@@

\begin{abstract}
  Расщепление или сплинт -- это разложение корневой системы в объединение нескольких корневых
  систем. Сплинт возникает при изучении (регулярных) вложений редуктивных подалгебр. Его можно
  использовать для построения правил ветвления представлений. Мы рассматриваем специальное вложения
  подалгебры Ли в простые алгебры Ли, классифицируем проекции корневых систем алгебр и выводим
  условия возникновения сплинта и совпадения коэффициентов ветвления с кратностями весов. Хотя
  такое совпадение происходит не слишком часто, оно оказывается связано с базисом Гельфанда-Цетлина.

\noindent{\it Keywords\/}: Алгебра Ли, специальная подалгебра, ветвление, кратность весов, сплинт
\end{abstract}

\section{Введение}
\label{sec:introduction}

Понятие сплинта (расщепления) было предложено Дэвидом Рихтером в работе \cite{richter2008splints}.
Сплинт -- это разложение корневой системы в объединение образов двух или более вложений других
корневых систем. Вложением  $\phi$ корневой системы $\Delta_1$ в корневую систему  system $\Delta$
называется биективное отображение корней $\Delta_{1}$ в (собственное) подмножество  $\Delta$,
коммутирующее с  законом сложения векторов в $\Delta_{1}$ и $\Delta$.
\begin{equation*}
\phi:\Delta_1 \longrightarrow \Delta
\end{equation*}
\begin{equation*}
\phi \circ (\alpha + \beta) =\phi \circ \alpha + \phi \circ \beta,
\,\,\, \alpha,\beta \in \Delta_1
\end{equation*}

Заметим, что мы не требуем от образа $Im(\phi)$ сохранения свойств корневой системы за исключением
совпадения правил сложения с правилами сложения в $\Delta_{1}$ (для прообразов). Если вложение
$\phi$ сохраняет углы между корнями, то оно называется ``метрическим''. Два вложения $\phi_1$ и
$\phi_2$ расщепляют корневую систему $\Delta$, если последняя может быть представлена в виде
дизъюнктного образов $Im(\phi_1)$ и $Im(\phi_2)$.

Чем же интересны отображения корневых систем, не сохраняющие углов между корнями? Аддитивные
свойства корневой системы определяют структуру модуля Верма:
    \begin{equation*}
      M^{\mu}=U(\gf)\underset{U(\bff_{+})}{\otimes} D^{\mu}(\bff_{+})=\{(E^{-\alpha_{1}})^{n_{1}}\dots (E^{-\alpha_{s}})^{n_{s}} \left|v_{\mu}\right>\}_{\alpha_{i}\in\Delta^{+}}^{n_{i}=0,1,\dots}
    \end{equation*}
Здесь $\bff_{+}$ -- подалгебра Бореля алгебры $\gf$, $D^{\mu}$ -- одномерное представление
подалгебры Бореля, $\left|v_{\mu}\right>$ -- вектор старшего веса, а  $E^{-\alpha_{j}}$ --
понижающие операторы, соответствующие положительным корням $\alpha_{j}\in \Delta^{+}$. 
Формула Вейля для характеров выражает характер неприводимого представления в виде комбинации характеров
модулей Верма: 
\begin{equation*}
  \mathrm{ch} L^{\mu}=\frac{\sum_{w\in W} \varepsilon(w) e^{w(\mu+\rho)-\rho}}{\sum_{w\in W}\varepsilon(w) e^{w\rho-\rho}}=\sum_{w\in W} \varepsilon(w)\; \mathrm{ch} M^{w(\mu+\rho)-\rho}
\end{equation*}
Здесь суммирование ведется по элементам группы Вейля, действие которых $w\triangleright \mu
=w(\mu+\rho)-\rho$ зависит от углов между корнями. Но формулу Вейля для характеров можно переписать,
представив элементы группы Вейля в виде произведений отражений $s_{\alpha}, \alpha \in S$ в
гиперплоскостях, ортогональных простым корням: $w=s_{\alpha_{1}}\cdot s_{\alpha_{2}}\dots$.
Отражения действуют на корневую систему перестановками, так что композицию такого действия с
вложением $\phi$ нетрудно вычислить. Орбиту группы Вейля $w\triangleright \mu, w\in W$ можно
построить последовательными вычитаниями корней из старшего веса $\mu$. Для этого обозначим индексы
Дынкина веса $\mu$ через $(\mu_{1},\dots \mu_{r})$. Тогда $\mu-\mu_{1}\alpha_{1},
\mu-\mu_{2}\alpha_{2},\dots, \mu-\mu_{r}\alpha_{r}, \alpha_{i}\in S$ -- это веса орбиты, которые
получаются из старшего веса $\mu$ элементарными отражениями $s_{\alpha_{i}}, \alpha_{i}\in S$.
Следующий набор точек орбиты, получаемых последовательным применением двух отражений
$s_{\alpha_{i}}\cdot s_{\alpha_{j}}\triangleright\mu$, можно построить путем вычитания
$\mu-\mu_{i}\alpha_{i}-\mu_{j} (s_{\alpha_{i}}\alpha_{j})$. Продолжая данную процедуру, мы построим
образ сингулярного элемента $\Psi=\sum_{w\in W} \varepsilon(w) e^{w(\mu+\rho)-\rho}$ под действием
вложения $\phi$ \cite{2011arXiv1111.6787L}. Кратности в характере модуля Верма не меняются при
вложении, так как они определяются аддитивными свойствами корней. Соответственно, при вложении
сохраняются и кратности в неприводимых модулях.

Корневая система регулярной подалгебры содержится в корневой системе алгебры, так что сплинт
возникает при вычислении коэффициентов ветвления.

В работе  \cite{2011arXiv1111.6787L} было показано, что существование сплинта ведет к совпадению
коэффициентов ветвления с кратностями весов при выполнении некоторых дополнительных условий.
Обозначим алгебру Ли через $\gf$ и рассмотрим ее подалгебру $\af$. Если подалгебра  $\af$ регулярна,
то ее корневая система  $\Delta_{\af}$ содержится в $\Delta_{\gf}$. Коэффициенты ветвления $b^{(\mu)}_{\nu}$
появляются при разложении неприводимого представления  $L^{\mu}_{\gf}$ алгебры $\gf$ в сумму
неприводимых представлений подалгебры $\af$:
\begin{equation}
  \label{eq:1}
  L^{\mu}_{\gf}=\bigoplus_{\nu} b^{(\mu)}_{\nu} L^{\nu}_{\af}
\end{equation}

Предположим, что корневая система $\Delta_{\gf}$ расщепляется как  $\Delta_{\gf}=\Delta_{\af} \cup
\phi(\Delta_{\sfr})$, где  $\phi$ -- вложение корневой системы $\Delta_{\sfr}$ некоторой полупростой
алгебры Ли $\sfr$. Тогда коэффициенты ветвления $b^{(\mu)}_{\nu}$ при редукции
$L^{(\mu)}_{\gf\downarrow \af}$ совпадают с кратностями весов  $m^{(\tilde \mu)}_{\nu}$ в
представлениях  $\sfr$ при выполнении определенного технического условия \cite{2011arXiv1111.6787L}.
Старший вес  $\tilde\mu$ представления $\sfr$ вычисляется по индексам Дынкина старшего веса  $\mu$
модуля алгебры $\gf$. Доказательство равенства коэффициентов ветвления с кратностями весов
представлений $\sfr$ основывается на разложении сингулярного элемента алгебры  $\gf$ в комбинацию
сингулярных элементов $\sfr$:
\begin{equation}
 \Phi ^{\mu }_{\gf} = \sum_{w\in W_{\frak{a}}}\epsilon\left( w\right)w\left( e^{\mu +\rho _{\gf}}\phi\left( e^{-\widetilde{\mu }}\Psi^{\widetilde{\mu }}_{\sfr}\right)\right),
\end{equation}
где $\epsilon\left( w\right)$ -- определитель элемента $w$ группы Вейля $W_{\af}$ подалгебры $\af$.
Действие группы Вейля на весах продолжается на алгебры формальных экспонент по правилу
$w\left(e^{\nu}\right)=e^{w\nu}, w\in W$. Отсюда кратность $M_{\left( \sfr\right) \widetilde{\nu
  }}^{\widetilde{\mu }}$ веса $\widetilde{\nu }$ из весовой диаграммы представления алгебры $\sfr$
со старшим весом $\tilde\mu$ определят коэффициент ветвления $b_{\nu }^{(\mu )}$ для старшего веса
$\nu =\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right) $:
\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{(\mu )}=M_{\left( \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}. 
\label{bran1}
\end{equation}

В данной работе мы воспользуемся похожим подходом для изучения специальных вложений. Мы
рассматриваем специальные вложения подалгебр Ли в алгебру Ли. В таком случае корневая система
подалгебры на содержится в корневой системе алгебры и первоначальная мотивация для введения сплинта
не применима. Но мы можем рассмотреть проекцию корневой системы алгебры на корневое пространство
подалгебры. Результат такой проекции уже не будет корневой системой, однако он будет удовлетворять
более мягким условиям (см Раздел \ref{sec:spec-embedd-proj}). Оказывается, что сплинты проекций
корневых систем можно классифицировать, используя разложение присоединенного представления алгебры
в прямую сумму представлений подалгебры. Мы показываем, что при наличии такого сплинта коэффициенты
ветвления совпадают с кратностями весов в представлениях другой алгебры Ли (Раздел
\ref{sec:splints-spec-embedd}). А значит, в этом случае коэффициенты ветвления очень легко вычислить.

Мы используем теорию представлений подалгебры для классификации всех сплинтов для проекции корневой
системы алгебры. Однако метод универсален и мы применяем его и к метрическим сплинтам и регулярным
подалгебрам. В результате мы получаем правила Гельфанда-Цетлина для регулярных и специальных
вложений.

В заключении \ref{sec:conclusion} мы обсуждаем случаи, которые не укладываются в упомянутую выше
классификацию.

\section{Специальные вложения и проекции корневой системы}
\label{sec:spec-embedd-proj}
